Online Multiclass Linear Classification with Bandit Feedback

### 2018-11-19 arxiv-version/main.tex
Add more explanations

% LaTeX packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fullpage}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{thmtools}
\usepackage{tikz}

@@

We study the \textsc{Online Multiclass Classification with Bandit Feedback}
problem. The protocol of the problem is stated below. The goal is to design an
algorithm that makes small number of mistakes $\sum_{t=1}^T z_t = \sum_{t=1}^T
\indicator{\widehat y_t \neq y_t}$. The problem was initially proposed by
\cite{Kakade-Shalev-Shwartz-Tewari-2008}. The corresponding problem with full
information feedback is solved by the classical \textsc{Multiclass Perceptron}
algorithm and its variants (see e.g.
\cite{Crammer-Dekel-Keshet-Shalev-Shwartz-Singer-2006}). The problem is quite
natural and potential applications include online advertising and online content
optimization where the feedback corresponds to a user's click.

@@

@@

The bandit feedback setting, however, is much more challenging. We start by
dissecting the notion of linear separability in the multiclass setting. Beside
the standard notion of linear separability, we define also and \emph{strong
linear separability}. For binary classification, these two definitions are
identical. However, they differ for multiclass classification. For multiclass
classification with $K$ classes the standard notion of linear separability means
that the examples from each class lie in an intersection of $K-1$ halfspaces and
the examples outside of the class lie in the complement of the intersection of
the halfspaces. Strong linear separability means that examples from each class
can be separated from the remaining examples by a \emph{single} hyperplane. We
state the definitions precisely in
Section~\ref{section:notions-of-linear-separability}.

For the case when the examples are strongly linearly separable, we design a
simple and efficient algorithm with expected number of mistakes at most $K
(R/\gamma)^2$. The algorithm can be viewed as running $K$ copies of the
\textsc{Binary Perceptron} algorithm, one copy for each class. Intuitively
speaking, the factor $K$ is the price we pay for the bandit feedback,
or more precisely, the lack of full-information feedback.

@@

Finally, we study two questions related to the computational and
information-theoretic hardness of the problem. Any algorithm for the bandit
setting collects information in the form of so called \emph{strongly labeled}
and \emph{weakly labeled} examples. Strongly-labeled examples are those for
which we know the class label. Weakly labeled example is an example for which we
know that class label can be anything except for a particular one class. First,
we show the offline problem of determining if a linear multiclass separator
exists for a set of strongly- and weakly-labeled examples is NP-complete. This
result is a simple consequence of the fact learning intersection of halfspaces
is NP-complete~\citep{Blum-Rivest-1993}. Second, we show that any algorithm that
uses only strongly-labeled examples and ignores weakly labeled examples makes at
least $\Omega(???)$ mistakes.

@@

\begin{definition}
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, and let $\gamma$ be a positive real number. Let $(x_1, y_1), (x_2,
y_2), \dots, (x_T, y_T)$ labeled examples in $V \times \{1,2,\dots,K\}$.

@@

\begin{figure}
\begin{center}
\input{figures/linearly-separable-examples-with-margin}
\end{center}
\caption{The figure shows labeled examples in $\R^2$. The examples belong to
$K=3$ classes colored white, gray and black respectively. Each class lies in a
$120^\circ$ wedge. While the examples are linearly separable with a positive
margin $\gamma$, they are \emph{not} strongly linearly separable for any
positive margin $\gamma$.
}
\label{figure:linearly-separable-examples-with-margin}
\end{figure}

@@

@@

\begin{theorem}
Let $(V, \ip{\cdot}{\cdot})$ be any inner product space, let $K$ be a positive
integer, $\gamma$ a positive real number, $R$ be a non-negative real number If
$(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled examples in
$V \times \{1,2,\dots,K\}$ such that the examples are strongly separable with
margin $\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ then the
expected number of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $K(K-1) \lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

Let $C$ be the number of times line 16 gets executed. Let $U$ be the number of
times line 16 or line 21 get executed. In other words, $U$ the number of times
the weight vectors are updated. Clearly, $U = A + C$. We know that $U \le
\frac{K} \lfloor (R/\gamma)^2 \rfloor$.

The key observation is that $\Exp[B] = (K-1) \Exp[C]$ since there the $1/K$
probability that the algorithm guesses the correct label correctly if $S_t =
\emptyset$. Putting all the information together, we get that
\begin{align*}
\Exp[M]
& = \Exp[A] + \Exp[B] \\
& = \Exp[A] + (K-1) \Exp[C]  \\
& \le (K-1) \Exp[A + C] \\
& = (K-1) \Exp[U]  \\
& \le K(K-1) \lfloor (R/\gamma)^2 \rfloor \; .
\end{align*}
\end{proof}

>>>>

% LaTeX packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{fullpage}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{thmtools}
\usepackage{tikz}

@@

We design new algorithms for the problem of \textsc{Online Multiclass
Classification with Bandit Feedback}. In this problem, in each round $t$, the
algorithm receives a feature vector $x_t$ and algorithm has to predict one of
$K$ possible classes. Imediately after the algorithm makes its prediction, it
receives feedback whether the prediction was correct or not. However, in
contrast with the standard, so called full information setting, the algorithm
does \emph{not} receive the correct label. In other words, the feedback is
purely binary (correct prediction or incorrect prediction). To allow for
meaningful randomized strategies, we assume that the correct label for round $t$
is chosen prior to algorithm's prediction in that round.\footnote{The reader can
assume throughout the paper a simpler model of the so called \emph{oblivious
adversary} where the whole sequence of labeled examples $(x_1, y_1), (x_2, y_2),
\dots, (x_T, y_T)$ is chosen before the first round of the game.} The protocol
of the problem is stated below.

@@

The goal is to design an algorithm that makes small number of mistakes
$\sum_{t=1}^T z_t = \sum_{t=1}^T \indicator{\widehat y_t \neq y_t}$. The problem
was initially proposed by \cite{Kakade-Shalev-Shwartz-Tewari-2008}. The
corresponding problem with full information feedback is solved by the classical
\textsc{Multiclass Perceptron} algorithm and its variants (see e.g.
\cite{Crammer-Dekel-Keshet-Shalev-Shwartz-Singer-2006}). The problem is quite
natural and potential applications include online advertising and online content
optimization where the feedback corresponds to a user's click.

@@

The bandit feedback setting, however, is much more challenging. To understand
the differences, we dissect the notion of linear separability in the multiclass
setting. Beside the standard notion of linear separability, we define also
\emph{strong linear separability}. For binary classification, these two
definitions are identical. However, they differ for multiclass classification.
For multiclass classification with $K$ classes the standard notion of linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside of the class lie in the complement of
the intersection of the halfspaces. Strong linear separability means that
examples from each class can be separated from the remaining examples by a
\emph{single} hyperplane. We state the definitions precisely in
Section~\ref{section:notions-of-linear-separability}.

For the case when the examples are strongly linearly separable, we design a
simple and efficient algorithm with expected number of mistakes at most $K(K-1)
\lfloor (R/\gamma)^2 \rfloor$. The algorithm can be viewed as running $K$ copies
of the \textsc{Binary Perceptron} algorithm, one copy for each class.
Intuitively speaking, the factor $K(K-1)$ is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback.

@@

Finally, we study two questions related to the computational and
information-theoretic hardness of the problem. Any algorithm for the bandit
setting collects information in the form of so called \emph{strongly labeled}
and \emph{weakly labeled} examples. Strongly-labeled examples are those for
which we know the class label. Weakly labeled example is an example for which we
know that class label can be anything except for a particular one class. First,
we show the offline problem of determining if a linear multiclass separator
exists for a set of strongly- and weakly-labeled examples is NP-complete. This
result is a simple consequence of the fact that learning intersection of
halfspaces is NP-complete~\citep{Blum-Rivest-1993}. Second, we show that any
algorithm that uses only strongly-labeled examples and ignores weakly labeled
examples makes at least $\Omega(???)$ mistakes.

@@

\begin{definition}
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, and let $\gamma$ be a positive real number. Let $(x_1, y_1), (x_2,
y_2), \dots, (x_T, y_T)$ be labeled examples in $V \times \{1,2,\dots,K\}$.

@@

\begin{figure}
\begin{center}
\input{figures/linearly-separable-examples-with-margin}
\end{center}
\caption[]{The figure shows labeled examples in $\R^2$. The examples belong to
$K=3$ classes colored white, gray and black respectively. Each class lies in a
$120^\circ$ wedge. In other words, each class lies in an intersection of two
halfspaces.

While the examples are linearly separable with a positive margin $\gamma$, they
are \emph{not} strongly linearly separable for any positive margin $\gamma$.
}
\label{figure:linearly-separable-examples-with-margin}
\end{figure}

@@

In this section, we present an algorithm for \textsc{Online Multiclass
Classification with Bandit Feedback} and prove an upper bound on the number of
mistakes under the assumption that the examples are strongly linearly separable
with a margin. The algorithm is stated
Algorithm~\eqref{algorithm:algorithm-for-strongly-linearly-separable-examples}
below and the mistake upper bound is stated as
\autoref{theorem:stronglys-separable-example-mistake-upper-bound}.

The idea behind the algorithm is to use $K$ copies of the \textsc{Binary
Perceptron} algorithm. Each copy corresponds to a class. In each round, each
copy makes a binary decision whether the feature vector belongs their respective
class or not. If at least one copy makes a positive prediction, the algorithm
predicts accordingly. If multiple copies make positive predictions, the
algorithm chooses one of them at random. If all copies make negative
predictions, the algorithms makes a blind guess.

@@

\begin{theorem}
\label{theorem:stronglys-separable-example-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be any inner product space, let $K$ be a positive
integer, $\gamma$ a positive real number, $R$ be a non-negative real number If
$(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled examples in
$V \times \{1,2,\dots,K\}$ such that the examples are strongly separable with
margin $\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ then the
expected number of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $K(K-1) \lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

Let $C$ be the number of times line 16 gets executed. Let $U$ be the number of
times line 16 or line 21 get executed. In other words, $U$ is the number of
times the weight vectors are updated. Clearly, $U = A + C$. We know that $U \le
K \lfloor (R/\gamma)^2 \rfloor$.

The key observation is that $\Exp[B] = (K-1) \Exp[C]$ since if $S_t =
\emptyset$, there is $1/K$ probability that the algorithm guesses the correct
label correctly and $(K-1)/K$ probability that it guesses is incorrectly.
Putting all the information together, we get that
\begin{align*}
\Exp[M]
& = \Exp[A] + \Exp[B] \\
& = \Exp[A] + (K-1) \Exp[C]  \\
& \le (K-1) \Exp[A + C] \\
& = (K-1) \Exp[U]  \\
& \le K(K-1) \lfloor (R/\gamma)^2 \rfloor \; .
\end{align*}
\end{proof}
